### 2010-10-24 Thesis/Chapter4.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter[Modelling the deformation of solid objects in real-time][Modelling solid objects in real-time]{Modelling the deformation of solid objects in real-time}
\label{chap4}
\begin{shortAbstract}
Modelling the deformation of solid organs is a crucial problem in medical simulation. If the finite element method is classicaly used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive. The main reason for this is that the finite element method is computationally demanding and not easily compatible with time constraints required in interactive simulation. Therefore, other methods were developed to allow faster computations, at the cost of larger approximations of course. Over time, along with the increase in computational power, methods became more elaborate and less approximative. These different approaches may be grouped into three main categories: (1) geometrically based techniques, (2) approaches physically motivated and (3) methods actually based on the equations of continuum mechanics. In this chapter, we will review the most classic methods applied to modelling of solid objects with a particular emphasis on medical simulation. 
\end{shortAbstract}

@@

Conversely, when medical simulation is applied to training, such a precision in terms of deformation is not always required. A visually realistic deformation is often sufficient for an inexperienced physician to learn the appropriate gestures required for a given medical procedure. However, in some cases, the feel of touch can be very important to master the procedure. For instance, during a colonoscopy procedure, force feedbacks give precious information about what type of loop is forming during the colonoscope's insertion and the physician can then react accordingly to prevent loop formations. Consequently, a good colonoscopy simulator is required to provide accurate force feedbacks. If the contact forces were not reproduced to the operator, he would never be able to learn how to detect loop forming. Obviously, the computation of a realistic force to be returned demands an appropriate mechanical modelling of all structures. In constrast of per-operative guidance, medical training demands real-time computations of the deformations.

@@

Because they do not derive from the equations of continuum mechanics, mass-spring models have a limited capacity to model the various aspects of a material like anisotropy, viscoelasticity etc. Even worse, according to \cite{Bourguignon00}, if all springs are set to the same stiffness, the mesh geometry may generate undesired anisotropy as shown in \fig{chap4:fig-undesiredAnisotropy}. However, some works tried to take advantage of this feature by giving directions of interest to their model by specifically designing the mesh to align springs on specific directions. For instance, it was used by \cite{Miller88} to animate the motion of snakes and worms. It was also used by \cite{Ng-Thow-Hing97} in their muscle model where some of the springs were aligned with the muscle fibers and more recently by \cite{Magjarevic09} for modelling the mechanical behaviour of the myocardial tissue. More details on controlling anisotropy in mass-spring systems may be found in \cite{Bourguignon00}. 
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=4cm]{chapter4/undesiredAnisotropy1.pdf}}
\hspace{0.2cm}
\subfloat[ ]{\includegraphics[width=4cm]{chapter4/undesiredAnisotropy2.pdf}}
\caption[Undesired anisotropy with mass-spring models]{The two meshes are undergoing a downward pull at their bottom. While the tetrahedral mass-spring system (a) shows signs of anisotropy, the hexahedral mesh (b) with springs aligned along the gravity does not. Image courtesy of \cite{Bourguignon00}. } 
\label{chap4:fig-undesiredAnisotropy}
\end{figure}

@@

Cutting is a problem for all techniques pre-computing the inverse of the stiffness matrix. Indeed, topological changes require updates of this matrix, which is too computationally heavy to be done in real-time. \cite{Nienhuys01} avoid the precomputations by approximating the inverse at runtime with an iterative algorithm (Conjugate Gradient). Moreover, instead of subdividing each tetrahedron to make the cut appear where the user performed it, they adapt the mesh locally so that there are always triangles on the scalpel path, and perform the cut along those triangles (see \fig{chap4:fig-cutUnderScalpel}). If the approach is interesting, the authors reported a lag between the scalpel and the realised cut. In addition, repositioning the nodes within the mesh can easily generate degeneracies, which have to be eliminated. In his PhD thesis, \cite{Lee10} presented techniques to update the precomputed inverse of the stiffness matrix rather than recomputing it, which is much more efficient. The proposed technique was also extended to integrate the concept of condensation, increasing computational efficiency and reducing memory requirements. 
%
\begin{figure}[ht]
\centering 
\subfloat[Surface selection]{\includegraphics[height=3cm]{chapter4/cutUnderScalpel1.pdf}}
\hspace{0.2cm}
\subfloat[Node snapping]{\includegraphics[height=3cm]{chapter4/cutUnderScalpel2.pdf}}
\hspace{0.2cm}
\subfloat[Dissect]{\includegraphics[height=3cm]{chapter4/cutUnderScalpel3.pdf}}
\caption[The three steps in performing a cut]{The three steps in performing a cut (represented here in 2D). Image courtesy of \cite{Nienhuys01}.} 
\label{chap4:fig-cutUnderScalpel}
\end{figure}

@@

As we know, the assumption of linearity is usually only valid for very small deformations and strains. If this is often an acceptable trade-off between the visual deformation result and the speed of the system, research was carried out to handle large deformations. However, since geometrically non-linear analyses are highly computationally demanding, another approach was found which allows large deformation while still relying on a linear strain measure: the \emph{co-rotational} methods \label{chap4:corotationalMethods}. A very complete description of the co-rotational approach is given by \cite{Felippa00}. As explained in section~\ref{chap2:strainMeasures}, an appropriate strain measure is a tensor which only measures the deformation applied to the object. This is the case of the Cauchy-Green or Green tensors for instance. Those strain metrics only measures deformation, they are invariant with respect to rigid-body transformations applied. However, geometrically linear analyses make use of the infinitesimal strain tensor, which is only an approximation of the Green tensor. Consequently, this linearised strain measure is not invariant with respect to rigid-body transformations. In particular, the error brought by the approximated (linear) strain measure increases with the rotation of the element. This error in the measure creates ghost forces which over-deforms the simulated object (see \fig{chap4:fig-cylinder}). The idea behind co-rotational methods is the decomposition of the motion into a rigid-body and deformational components. In other words, the deformation of each element may be seen as measured in a coordinate system which follows the rigid-body motion of the element, hence cancelling the part of the error due to the rotation in the strain measure. Consequently, co-rotational formulations allow large displacements and rotations (since the rigid-body motion is not considered) as long as the strain remains small (the strain measure is still linear). 
%
\begin{figure}[ht]
\centering 
\includegraphics[height=6cm]{chapter4/cylinder1.pdf}
\hspace{0.5cm}
\includegraphics[height=6cm]{chapter4/cylinder2.pdf}
\caption[Deformation of cylinders]{A geometrically linear finite element model (green) is compared to a co-rotational finite element model (yellow). While still using a simple linear strain measure, the co-rotational formulation is not sensitive to rotations and thus does not overdeform.}
\label{chap4:fig-cylinder}
\end{figure}

@@

To the best of our knowledge, \cite{Rumpf01} were the firsts to leverage the power of GPUs to solve partial differential equations. They established a correspondence between common mathematical operations and OpenGL operations and structures. Thus, as an example, a vector is represented as a RGBA color vector in a color image on the graphics card. They decided to implement an iterative solver for a linear system of equations, which we find at the core of most finite element codes. Using this hardware accelerated solver, they solved the linear heat equations and presented an anisotropic diffusion model for image processing. They obtained promising results but noted the hardware restrictions that forced them to approximate all non-linear functions by linear ones in the implementation of the anisotropic diffusion. Although a complete hardware accelerated finite element method implementation was yet to be done, \citeauthor{Rumpf01} definitely stepped towards the right direction.

@@

A meshless method was introduced for the first time for modelling soft tissue in medical simulation by \cite{De01}. They use the so-called method of finite spheres (MFS). Nodal points are only sprinkled locally around the surgical tool tip and not over the whole domain. The interpolation is performed by functions that are nonzero only on spheres surrounding the nodes. A point collocation technique is used to generate the discrete equations. The authors qualitatively compared the method of finite spheres with the finite element technique from a commercial package and the displacement profile was similar in both cases. They were able to achieve a computational rate of about $ 100\, $Hz when 34 spheres where used. Real-time haptic rates of about $ 1\, $kHz was then obtained using a force extrapolation technique.

@@

Meshless methods were then introduced to computer graphics by \cite{Muller04}. In each step, they compute the spatial derivatives of the displacement field using a moving least squares (MLS) procedure. From these derivatives, they obtain strains, stresses and elastic forces at each simulated point. In addition, they proposed techniques for modelling and animating a point-sampled surface that dynamically adapts to deformations of the underlying volumetric model. They used a non-linear measure of strain with a linear constitutive model. They reported real-time simulation of models exibiting elastic, plastic, melting and flowing effects.

@@

The main advantage of meshless methods is of course in the absence of a mesh. This advantage of not having to go through the complex process of meshing was often claimed by meshless methods partisans. It also allows much easier cutting as it saves the cost of a computationally demanding remeshing around the cut. Yet, in contrast of finite element methods, meshless techniques are young and still in development. Boundary conditions are an aspect of the numerical solution of partial equations where meshless methods have many difficulties for instance. According to \cite{Horton10}, if meshless methods are accurate in terms of overall reaction forces, they are not as good with individual displacements or forces. Therefore the authors suggest to use a meshless algorithm to fill the interior sections of an organ (where a local discrepancy has minor consequences) and use the finite element method for the boundary. As a conclusion, if meshless methods are promising, the technique does not seem as mature as the finite element method for modelling soft tissues.

>>>>

\chapter{Modelling the deformation of solid objects in real-time}
\label{chap4}
\begin{shortAbstract}
Modelling the deformation of solid organs is a crucial problem in medical simulation. If the finite element method is classically used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive. The main reason for this is that the finite element method is computationally demanding and not easily compatible with time constraints required in interactive simulation. Therefore, other methods were developed to allow faster computations, at the cost of larger approximations of course. Over time, along with the increase in computational power, methods became more elaborate and less approximative. These different approaches may be grouped into three main categories: (1) geometrically based techniques, (2) approaches physically motivated and (3) methods actually based on the equations of continuum mechanics. In this chapter, we will review the most classic methods applied to modelling of solid objects with a particular emphasis on medical simulation. 
\end{shortAbstract}

@@

Conversely, when medical simulation is applied to training, such a precision in terms of deformation is not always required. A visually realistic deformation is often sufficient for an inexperienced physician to learn the appropriate gestures required for a given medical procedure. However, in some cases, the feel of touch can be very important to master the procedure. For instance, during a colonoscopy procedure, force feedbacks give precious information about what type of loop is forming during the colonoscope's insertion and the physician can then react accordingly to prevent loop formations. Consequently, a good colonoscopy simulator is required to provide accurate force feedbacks. If the contact forces were not reproduced to the operator, he would never be able to learn how to detect loop forming. Obviously, the computation of a realistic force to be returned demands an appropriate mechanical modelling of all structures. In contrast of per-operative guidance, medical training demands real-time computations of the deformations.

@@

Because they do not derive from the equations of continuum mechanics, mass-spring models have a limited capacity to model the various aspects of a material like anisotropy, viscoelasticity etc. Even worse, according to \cite{Bourguignon00}, if all springs are set to the same stiffness, the mesh geometry may generate undesired anisotropy as shown in \fig{chap4:fig-undesiredAnisotropy}. However, some works tried to take advantage of this feature by giving directions of interest to their model by specifically designing the mesh to align springs on specific directions. For instance, it was used by \cite{Miller88} to animate the motion of snakes and worms. It was also used by \cite{Ng-Thow-Hing97} in their muscle model where some of the springs were aligned with the muscle fibres and more recently by \cite{Magjarevic09} for modelling the mechanical behaviour of the myocardial tissue. More details on controlling anisotropy in mass-spring systems may be found in \cite{Bourguignon00}. 
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=4cm]{chapter4/undesiredAnisotropy1.pdf}}
\hspace{0.2cm}
\subfloat[ ]{\includegraphics[width=4cm]{chapter4/undesiredAnisotropy2.pdf}}
\caption[Undesired anisotropy with mass-spring models]{The two meshes are undergoing a downward pull at their bottom. While the tetrahedral mass-spring system (a) shows signs of anisotropy, the hexahedral mesh (b) with springs aligned along the gravity does not. Image courtesy of \cite{Bourguignon00}. } 
\label{chap4:fig-undesiredAnisotropy}
\end{figure}

@@

Cutting is a problem for all techniques pre-computing the inverse of the stiffness matrix. Indeed, topological changes require updates of this matrix, which is too computationally heavy to be done in real-time. \cite{Nienhuys01} avoid the pre-computations by approximating the inverse at runtime with an iterative algorithm (Conjugate Gradient). Moreover, instead of subdividing each tetrahedron to make the cut appear where the user performed it, they adapt the mesh locally so that there are always triangles on the scalpel path, and perform the cut along those triangles (see \fig{chap4:fig-cutUnderScalpel}). If the approach is interesting, the authors reported a lag between the scalpel and the realised cut. In addition, repositioning the nodes within the mesh can easily generate degeneracies, which have to be eliminated. In his PhD thesis, \cite{Lee10} presented techniques to update the precomputed inverse of the stiffness matrix rather than recomputing it, which is much more efficient. The proposed technique was also extended to integrate the concept of condensation, increasing computational efficiency and reducing memory requirements. 
%
\begin{figure}[ht]
\centering 
\subfloat[Surface selection]{\includegraphics[height=3cm]{chapter4/cutUnderScalpel1.pdf}}
\hspace{0.2cm}
\subfloat[Node snapping]{\includegraphics[height=3cm]{chapter4/cutUnderScalpel2.pdf}}
\hspace{0.2cm}
\subfloat[Dissect]{\includegraphics[height=3cm]{chapter4/cutUnderScalpel3.pdf}}
\caption[The three steps in performing a cut]{The three steps in performing a cut (represented here in 2D). Image courtesy of \cite{Nienhuys01}.} 
\label{chap4:fig-cutUnderScalpel}
\end{figure}

@@

As we know, the assumption of linearity is usually only valid for very small deformations and strains. If this is often an acceptable trade-off between the visual deformation result and the speed of the system, research was carried out to handle large deformations. However, since geometrically non-linear analyses are highly computationally demanding, another approach was found which allows large deformation while still relying on a linear strain measure: the \emph{co-rotational} methods \label{chap4:corotationalMethods}. A very complete description of the co-rotational approach is given by \cite{Felippa00}. As explained in section~\ref{chap2:strainMeasures}, an appropriate strain measure is a tensor which only measures the deformation applied to the object. This is the case of the Cauchy-Green or Green tensors for instance. Those strain metrics only measures deformation, they are invariant with respect to rigid-body transformations applied. However, geometrically linear analyses make use of the infinitesimal strain tensor, which is only an approximation of the Green tensor. Consequently, this linearised strain measure is not invariant with respect to rigid-body transformations. In particular, the error brought by the approximated (linear) strain measure increases with the rotation of the element. This error in the measure creates ghost forces which over-deforms the simulated object (see \fig{chap4:fig-cylinder}). The idea behind co-rotational methods is the decomposition of the motion into a rigid-body and deformational components. In other words, the deformation of each element may be seen as measured in a coordinate system which follows the rigid-body motion of the element, hence cancelling the part of the error due to the rotation in the strain measure. Consequently, co-rotational formulations allow large displacements and rotations (since the rigid-body motion is not considered) as long as the strain remains small (the strain measure is still linear). 
%
\begin{figure}[ht]
\centering 
\includegraphics[height=6cm]{chapter4/cylinder1.pdf}
\hspace{0.5cm}
\includegraphics[height=6cm]{chapter4/cylinder2.pdf}
\caption[Deformation of cylinders]{A geometrically linear finite element model (green) is compared to a co-rotational finite element model (yellow). While still using a simple linear strain measure, the co-rotational formulation is not sensitive to rotations and thus does not over-deform.}
\label{chap4:fig-cylinder}
\end{figure}

@@

To the best of our knowledge, \cite{Rumpf01} were the first ones to leverage the power of GPUs to solve partial differential equations. They established a correspondence between common mathematical operations and OpenGL operations and structures. Thus, as an example, a vector is represented as a RGBA color vector in a color image on the graphics card. They decided to implement an iterative solver for a linear system of equations, which we find at the core of most finite element codes. Using this hardware accelerated solver, they solved the linear heat equations and presented an anisotropic diffusion model for image processing. They obtained promising results but noted the hardware restrictions that forced them to approximate all non-linear functions by linear ones in the implementation of the anisotropic diffusion. Although a complete hardware accelerated finite element method implementation was yet to be done, \citeauthor{Rumpf01} definitely stepped towards the right direction.

@@

A meshless method was introduced for the first time for modelling soft tissue in medical simulation by \cite{De01}. They use the so-called method of finite spheres (MFS). Nodal points are only sprinkled locally around the surgical tool tip and not over the whole domain. The interpolation is performed by functions that are non-zero only on spheres surrounding the nodes. A point collocation technique is used to generate the discrete equations. The authors qualitatively compared the method of finite spheres with the finite element technique from a commercial package and the displacement profile was similar in both cases. They were able to achieve a computational rate of about $ 100\, $Hz when 34 spheres where used. Real-time haptic rates of about $ 1\, $kHz was then obtained using a force extrapolation technique.

@@

Meshless methods were then introduced to computer graphics by \cite{Muller04}. In each step, they compute the spatial derivatives of the displacement field using a moving least squares (MLS) procedure. From these derivatives, they obtain strains, stresses and elastic forces at each simulated point. In addition, they proposed techniques for modelling and animating a point-sampled surface that dynamically adapts to deformations of the underlying volumetric model. They used a non-linear measure of strain with a linear constitutive model. They reported real-time simulation of models exhibiting elastic, plastic, melting and flowing effects.

@@

The main advantage of meshless methods is of course in the absence of a mesh. This advantage of not having to go through the complex process of meshing was often claimed by meshless methods partisans. It also allows much easier cutting as it saves the cost of a computationally demanding re-meshing around the cut. Yet, in contrast of finite element methods, meshless techniques are young and still in development. Boundary conditions are an aspect of the numerical solution of partial equations where meshless methods have many difficulties for instance. According to \cite{Horton10}, if meshless methods are accurate in terms of overall reaction forces, they are not as good with individual displacements or forces. Therefore the authors suggest to use a meshless algorithm to fill the interior sections of an organ (where a local discrepancy has minor consequences) and use the finite element method for the boundary. As a conclusion, if meshless methods are promising, the technique does not seem as mature as the finite element method for modelling soft tissues.

\section{Conclusion}

Modelling the deformation of solid organs is a crucial problem in medical simulation. If the finite element method is classically used in mechanical engineering for its ability to solve the equations of continuum mechanics fairly accurately, its development in medical simulation was more progressive. The main reason for this is that the finite element method is computationally demanding and not easily compatible with time constraints required in interactive simulation. Therefore, other methods were developed to allow faster computations, at the cost of larger approximations of course. Following the state of the art presented in this chapter, we can highlight a few trends in medical simulation.

Firstly, over time (along with the increase in computational power) models used to describe anatomical structures became more elaborate and less approximative. Medical simulators become more concerned with the accuracy of their deformation. While they used to only cover a subset of their actual mechanical behaviour, simulators now began to reproduce most aspects of tissue behaviour under various types of loading (non-linearity, anisotropy and viscoelasticity) as described by \cite{Fung93}.

Secondly, we observe a tendency towards implementations on parallel architectures, and GPUs in particular. Despite limited hardware possibilities, GPUs allow speedups up to orders of magnitude. Of course, such a gain is not straightforwardly obtained and often necessitates a substantial re-write of the algorithm to achieve an optimal result. But in environments with real-time constraints such as medical simulators, hardware acceleration through the use of GPUs is definitely an interesting path to follow.

At last, we can identify two types of anatomical structures in human body. They can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.). Despite this assertion, it is worth mentioning that all the models presented so far were applied to solid organ modelling. Indeed, in the medical simulation community, much more attention is allotted to solid organs and the modelling of this type of structure is better understood.

We recall that one of the challenge faced by the field of medical simulation is to model the deformation of organs in a physically realistic manner. In addition, this objective has to be achieved within heavy time constraints. To answer this problem for solid organs, our idea is to combine the two recent trends observed in medical simulation: more elaborate biomechanical models and GPU acceleration. Our contributions towards this issue will be presented in the two following chapters. Since specific approaches are required for simulating the deformation of thin anatomical structures, this problem will be tackled from chapter~\ref{chap7}.

### 2010-10-24 Thesis/Chapter5.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Some important observations arise from \eqref{chap5:criticalTimestep} and \eqref{chap5:waveSpeed}. Firstly, since the stiffness of soft tissues is very much smaller than that of common engineering materials ($E \approx 3 \times 10^3\, $Pa for brain tissue versus $\approx 2 \times 10^{11}\, $Pa for steel), the allowable time step is much larger for analysis of the former. This is a key reason for the expediency of explicit time integration for analysis of soft tissues. Secondly, soft tissues are generally considered to be incompressible, leading to $\nu \approx 0.49$ commonly being employed. However, if this value can be relaxed even further, significant increases in $ \Delta t_{cr}$ are possible. As an example, for a given $ L_e $ a material with $E = 3\,000\, $Pa and $\nu = 0.45$ allows a time step of more than double that of a material with $\nu = 0.49$. Of course, lowering the Poisson ratio introduces inaccuracies, but for some applications (interactive medical simulation for instance) it may be that this is acceptable. Finally, it must be highlighted that $ \Delta t_{cr}$ is proportional to $ L_e $. If the element size is decreased (by using a finer mesh, or by analysing very small objects), inexorably $ \Delta t_{cr}$ is also reduced.

@@

Anisotropic mechanical response may arise, for example, from the presence of a highly organised microstructure such as those of connective tissues. These are predominantly composed of collagen or elastin fibres embedded in an amorphous matrix \citep{Fung93}, and may be considered as fibre reinforced composites in some cases. Alternatively, the presence of vasculature and other functional components means even non-load bearing organs may exhibit directional dependence \citep{Picinbono01,Prange02}. Whereas isotropic constitutive models may be formulated in terms of the usual principal strain invariants (as used in \eqref{chap5:strainEnergyFunction} for a neo-Hookean material), directional dependence requires inclusion of so-called pseudo-invariants of strain and material direction.

In order to model such soft-tissue features with the explicit TLED formulation without significant performance penalties, an efficient constitutive update procedure involving time integration of the relevant hereditary equations is required. We address this problem by presenting a procedure similar to that developed by \cite{Poon98} for analysis of anisotropic linear viscoelastic models. This was also adapted by \cite{Taylor07} for solution of their fibre composite-based microstructural model. In these cases the rate independent responses were based on linear elasticity, rather than a hyperelastic formulation suitable for large deformations. In the present work we begin with a class of anisotropic visco-hyperelastic models, and develop a constitutive update procedure for explicit analyses based on these.

>>>>

Some important observations arise from \eqref{chap5:criticalTimestep} and \eqref{chap5:waveSpeed}. Firstly, since the stiffness of soft tissues is very much smaller than that of common engineering materials ($E \approx 3 \times 10^3\, $Pa for brain tissue versus $\approx 2 \times 10^{11}\, $Pa for steel), the allowable time step is much larger for analysis of the former. This is a key reason for the expediency of explicit time integration for analysis of soft tissues. Secondly, soft tissues are generally considered to be incompressible, leading to $\nu \approx 0.49$ commonly being employed. However, if this value can be relaxed even further, significant increases in $ \Delta t_{cr}$ are possible. As an example, for a given $ L_e $ a material with $E = 3\,000\, $Pa and $\nu = 0.45$ allows a time step of more than double that of a material with $\nu = 0.49$. Of course, lowering the Poisson's ratio introduces inaccuracies, but for some applications (interactive medical simulation for instance) it may be that this is acceptable. Finally, it must be highlighted that $ \Delta t_{cr}$ is proportional to $ L_e $. If the element size is decreased (by using a finer mesh, or by analysing very small objects), inexorably $ \Delta t_{cr}$ is also reduced.

@@

Anisotropic mechanical response may arise, for example, from the presence of a highly organised micro-structure such as those of connective tissues. These are predominantly composed of collagen or elastin fibres embedded in an amorphous matrix \citep{Fung93}, and may be considered as fibre reinforced composites in some cases. Alternatively, the presence of vasculature and other functional components means even non-load bearing organs may exhibit directional dependence \citep{Picinbono01,Prange02}. Whereas isotropic constitutive models may be formulated in terms of the usual principal strain invariants (as used in \eqref{chap5:strainEnergyFunction} for a neo-Hookean material), directional dependence requires inclusion of so-called pseudo-invariants of strain and material direction.

In order to model such soft-tissue features with the explicit TLED formulation without significant performance penalties, an efficient constitutive update procedure involving time integration of the relevant hereditary equations is required. We address this problem by presenting a procedure similar to that developed by \cite{Poon98} for analysis of anisotropic linear viscoelastic models. This was also adapted by \cite{Taylor07c} for solution of their fibre composite-based micro-structural model. In these cases the rate independent responses were based on linear elasticity, rather than a hyperelastic formulation suitable for large deformations. In the present work we begin with a class of anisotropic visco-hyperelastic models, and develop a constitutive update procedure for explicit analyses based on these.
